---
title: "00b_NA_management_and_imputation"
author: "Adriana Ibáñez"
date: "2025-12-01"
output: html_document
---

# *Summary*
This script performs a detailed analysis of missing data (NA) in the _Rancho Bernardo_ dataset for visits 7 to 10. The idea is to handle missing values by first identifying variables with a high percentage of missing data (more than 60% NA), and delete them from the db.
Finally, preparing the dataset for K-Nearest Neighbors (KNN) imputation separately for men and women datasets to have a final product with imputed values ( _db_final_female_ and _db_final_male_)


````{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)    # Manipulación de datos
library(tibble)
library(dplyr)
library(tidyr)
library(ggplot2)
library(magrittr)      # Para usar %>%
library(knitr)         # Para kable()
library(FactoMineR)
library(factoextra)
library(VIM)
library(openxlsx)
library(missForest)
library(VIM)    # Para KNN
library(mice)   # Para PMM
library(yaImpute) # Para NNI
library(randomForest)
library(kableExtra)
library(class)
library(caret)
````



### *PRE- ANALYSIS: VARIABLES WITH <50% DATA IN VISITS 7-10*
Delete the variables that have more than 50% of NA in each dataset (men and women).


```{r}
db_visit7 <- readRDS("db_visit7.rds")
db_visit8 <- readRDS("db_visit8.rds")
db_visit9 <- readRDS("db_visit9.rds")
db_visit10 <- readRDS("db_visit10.rds")

db_filt <- readRDS("db_7_to_10.rds")
```


```{r}
# Function: to compare NA per variable and add visit name
na_summary <- function(df, visita) {
  tibble(
    variable = names(df),
    na_count = sapply(df, function(x) sum(is.na(x))),
    visit = visita
  )
}

# Create individual table
na7  <- na_summary(db_visit7,  "visit7")
na8  <- na_summary(db_visit8,  "visit8")
na9  <- na_summary(db_visit9,  "visit9")
na10 <- na_summary(db_visit10, "visit10")

# Join the tables
na_combined <- bind_rows(na7, na8, na9, na10)

# Wide format
na_table <- na_combined %>%
  pivot_wider(names_from = visit, values_from = na_count, values_fill = 0) %>%
  arrange(desc(visit7))  # Sort by visit 7 first

print(na_table)
```


#### General NA analysis
```{r}
# Calculate total N per visit (all subjects together)
n_total_v7 <- nrow(db_visit7)
n_total_v8 <- nrow(db_visit8) 
n_total_v9 <- nrow(db_visit9)
n_total_v10 <- nrow(db_visit10)

# Añadir % disponibles a na_combined
completitud_710 <- na_combined %>%
  mutate(
    n_total = case_when(
      visit == "visit7" ~ n_total_v7,
      visit == "visit8" ~ n_total_v8,
      visit == "visit9" ~ n_total_v9,
      visit == "visit10" ~ n_total_v10
    ),
    pct_disponibles = round((1 - na_count/n_total) * 100, 1)
  ) %>%
  select(variable, visit, n_total, na_count, pct_disponibles)

# CALCULAR % NA EXACTO: suma(visit7+visit8+visit9+visit10) / (590 × 4)
n_total_visitas <- 590 * 4  # 2360 observaciones totales esperadas

completitud_exacta <- na_table %>%
  mutate(
    na_total = visit7 + visit8 + visit9 + visit10,  # Suma NA de las 4 visitas
    datos_disponibles = n_total_visitas - na_total,  # Datos NO NA
    pct_na = round((na_total / n_total_visitas) * 100, 1),
    pct_disponibles = round((datos_disponibles / n_total_visitas) * 100, 1)
  ) %>%
  select(variable, na_total, datos_disponibles, pct_na, pct_disponibles) %>%
  arrange(pct_disponibles)

# VARIABLES CON <50% DATOS (CRITERIO CONSERVADOR)
variables_bajas <- completitud_exacta %>%
  filter(pct_na > 60) %>%
  arrange(desc(pct_na))


# EXPORTAR CSVs
write.csv(completitud_exacta, "output/completitud_variables_7_10.csv", row.names = FALSE)
write.csv(variables_bajas, "output/variables_menos_50pct_7_10.csv", row.names = FALSE)

```

### *Separation by men and women + individual KNN*

First, we separate the dataset into men and women:
```{r}
db_female <- db_filt %>% filter(SEX == "0")
db_male   <- db_filt %>% filter(SEX == "1")
```

```{r}
mean(db_female$HT_IN)
mean(db_male$HT_IN)
```
```{r}
length(unique(db_female$SUBJID))
length(unique(db_male$SUBJID))
```



### _NA to more than 50% in WOMEN_
```{r}
# Create na_table WOMEN
na7_f <- na_summary(db_female %>% filter(VISIT == 7), "visit7")
na8_f <- na_summary(db_female %>% filter(VISIT == 8), "visit8")
na9_f <- na_summary(db_female %>% filter(VISIT == 9), "visit9")
na10_f <- na_summary(db_female %>% filter(VISIT == 10), "visit10")

na_combined_f <- bind_rows(na7_f, na8_f, na9_f, na10_f)
na_table_f <- na_combined_f %>%
  pivot_wider(names_from = visit, values_from = na_count, values_fill = 0)

# Calculate total N per visit WOMEN
n_total_v7_f <- nrow(db_female %>% filter(VISIT == 7))
n_total_v8_f <- nrow(db_female %>% filter(VISIT == 8))
n_total_v9_f <- nrow(db_female %>% filter(VISIT == 9))
n_total_v10_f <- nrow(db_female %>% filter(VISIT == 10))

# % available PER VISIT WOMEN
completitud_710_f <- na_combined_f %>%
  mutate(
    n_total = case_when(
      visit == "visit7" ~ n_total_v7_f,
      visit == "visit8" ~ n_total_v8_f,
      visit == "visit9" ~ n_total_v9_f,
      visit == "visit10" ~ n_total_v10_f
    ),
    pct_disponibles = round((1 - na_count/n_total) * 100, 1)
  ) %>%
  select(variable, visit, n_total, na_count, pct_disponibles)

# % EXACT NA WOMEN: sum NA / (368 × 4)
n_total_visitas_f <- 368 * 4  # 1472 observations
completitud_exacta_f <- na_table_f %>%
  mutate(
    na_total = visit7 + visit8 + visit9 + visit10,
    datos_disponibles = n_total_visitas_f - na_total,
    pct_na = round((na_total / n_total_visitas_f) * 100, 1),
    pct_disponibles = round((datos_disponibles / n_total_visitas_f) * 100, 1)
  ) %>%
  select(variable, na_total, datos_disponibles, pct_na, pct_disponibles) %>%
  arrange(pct_disponibles)

# VARIABLES with >60% NA WOMEN
variables_bajas_f <- completitud_exacta_f %>%
  filter(pct_na > 60) %>%
  arrange(desc(pct_na))


completitud_exacta_f %>% 
  filter(pct_na > 60) %>% 
  pull(variable)

# EXPORT
write.csv(completitud_exacta_f, "output/completitud_female_7_10.csv", row.names = FALSE)
write.csv(variables_bajas_f, "output/variables_mas_60pct_NA_female.csv", row.names = FALSE)
```


#### Delete this variables
```{r}
vars_f_high_na <- c("TESTOSTERONE", "ESTRADIOL", "BIO_TESTOSTERONE", "BIO_ESTRADIOL", 
                     "ALBUMIN_DIPSTICK", "INSULIN_CHALLENGE", "GHBHPLC", "PROINSULIN", 
                     "GLUCOSE_CHALLENGE", "REDCOUNT", "HEMOGLOBIN", "HEMATOCRIT", 
                     "WHITECOUNT", "MCV", "MCH", "MCHC", "CPEPTIDE", "PINP", 
                     "NTX_SERUM", "NTX_URINE_VITROS", "CRP", "GGT", "LDH", 
                     "URIC_ACID")

db_female <- db_female %>%
  select(-all_of(vars_f_high_na))
```


### _NA more than 60% MEN_

```{r}
# Create na_table MEN 
na7_m <- na_summary(db_male %>% filter(VISIT == 7), "visit7")
na8_m <- na_summary(db_male %>% filter(VISIT == 8), "visit8")
na9_m <- na_summary(db_male %>% filter(VISIT == 9), "visit9")
na10_m <- na_summary(db_male %>% filter(VISIT == 10), "visit10")

na_combined_m <- bind_rows(na7_m, na8_m, na9_m, na10_m)
na_table_m <- na_combined_m %>%
  pivot_wider(names_from = visit, values_from = na_count, values_fill = 0)

# Calculate total N per visit MEN
n_total_v7_m <- nrow(db_male %>% filter(VISIT == 7))
n_total_v8_m <- nrow(db_male %>% filter(VISIT == 8))
n_total_v9_m <- nrow(db_male %>% filter(VISIT == 9))
n_total_v10_m <- nrow(db_male %>% filter(VISIT == 10))

# % available PER VISIT MEN
completitud_710_m <- na_combined_m %>%
  mutate(
    n_total = case_when(
      visit == "visit7" ~ n_total_v7_m,
      visit == "visit8" ~ n_total_v8_m,
      visit == "visit9" ~ n_total_v9_m,
      visit == "visit10" ~ n_total_v10_m
    ),
    pct_disponibles = round((1 - na_count/n_total) * 100, 1)
  ) %>%
  select(variable, visit, n_total, na_count, pct_disponibles)

# % NA EXACT MEN: sum NA / (222 × 4)
n_total_visitas_m <- 222 * 4  # 888 observations
completitud_exacta_m <- na_table_m %>%
  mutate(
    na_total = visit7 + visit8 + visit9 + visit10,
    datos_disponibles = n_total_visitas_m - na_total,
    pct_na = round((na_total / n_total_visitas_m) * 100, 1),
    pct_disponibles = round((datos_disponibles / n_total_visitas_m) * 100, 1)
  ) %>%
  select(variable, na_total, datos_disponibles, pct_na, pct_disponibles) %>%
  arrange(pct_disponibles)

# VARIABLES WITH >60% NA MEN
variables_bajas_m <- completitud_exacta_m %>%
  filter(pct_na > 60) %>%
  arrange(desc(pct_na))


completitud_exacta_m %>% 
  filter(pct_na > 60) %>% 
  pull(variable)


# EXPORT
write.csv(completitud_exacta_m, "output/completitud_male_7_10.csv", row.names = FALSE)
write.csv(variables_bajas_m, "output/variables_mas_60pct_NA_male.csv", row.names = FALSE)
```

##### Delete this variables in MEN db
```{r}
vars_m_high_na <- c("ALBUMIN_DIPSTICK", "INSULIN_CHALLENGE", "BIO_ESTRADIOL", 
                     "ESTRADIOL", "BIO_TESTOSTERONE", "TESTOSTERONE", "GHBHPLC", 
                     "PROINSULIN", "GLUCOSE_CHALLENGE", "HEMOGLOBIN", "HEMATOCRIT", 
                     "WHITECOUNT", "REDCOUNT", "MCV", "MCH", "MCHC", "CPEPTIDE", 
                     "NTX_URINE_VITROS", "CRP", "PINP", "NTX_SERUM", "GGT", 
                     "LDH", "URIC_ACID")

db_male <- db_male %>%
  select(-all_of(vars_m_high_na))

```

### *KNN performing*

```{r}
# Remove non-imputable variables
db_knn_female <- db_female %>%
  select(-DEATHAGE, -EVERSMOKE, -HXSMOKE, -PASTALC, -ALCFREQ, -EX3XWK, -REGEX, -EX_COMP10YRS, -SLEEPING_HRS, -NAPPING_HRS, -SITTING_HRS)

db_knn_male <- db_male %>%
  select(-DEATHAGE, -EVERSMOKE, -HXSMOKE, -PASTALC, -ALCFREQ, -EX3XWK, -REGEX, -EX_COMP10YRS, -SLEEPING_HRS, -NAPPING_HRS, -SITTING_HRS)
```

#### KNN preparation: remove SUBJID and VISIT, select numeric variables only
```{r}
# Women
f_split <- db_knn_female %>%
  select(-SUBJID, -VISIT) %>%
  select(where(is.numeric))

f_nonnum <- db_knn_female %>%
  select(SUBJID, VISIT)  # Save IDs

# Men
m_split <- db_knn_male %>%
  select(-SUBJID, -VISIT) %>%
  select(where(is.numeric))

m_nonnum <- db_knn_male %>%
  select(SUBJID, VISIT)

## Simple scaling before KNN (to avoid bias in distances)
# Women: Calculate means and standard deviations ignoring NA
f_means <- colMeans(f_split, na.rm = TRUE)
f_sds <- apply(f_split, 2, sd, na.rm = TRUE)
f_sds[f_sds == 0] <- 1  # Avoid division by zero if variable is constant

f_scaled <- sweep(f_split, 2, f_means, "-")  # - average
f_scaled <- sweep(f_scaled, 2, f_sds, "/")   # divide for sd
```

```{r}
# Men
m_means <- colMeans(m_split, na.rm = TRUE)
m_sds <- apply(m_split, 2, sd, na.rm = TRUE)
m_sds[m_sds == 0] <- 1

m_scaled <- sweep(m_split, 2, m_means, "-")
m_scaled <- sweep(m_scaled, 2, m_sds, "/")

## KNN en datos escalados
f_knn_scaled <- kNN(f_scaled, k = 10) %>%
  select(-ends_with("_imp"))

m_knn_scaled <- kNN(m_scaled, k = 10) %>%
  select(-ends_with("_imp"))

## Desescalado simple
f_knn <- sweep(f_knn_scaled, 2, f_sds, "*")  # Multiply for sd
f_knn <- sweep(f_knn, 2, f_means, "+")      # + media

m_knn <- sweep(m_knn_scaled, 2, m_sds, "*")
m_knn <- sweep(m_knn, 2, m_means, "+")
```

#### Chekk remaining NAs
```{r}
## Quick check of remaining NA (to debug your error)
cat("NA restantes en mujeres después de imputar:", sum(is.na(f_knn)), "\n")
cat("NA restantes en hombres después de imputar:", sum(is.na(m_knn)), "\n")

```

#### Reconstruct final datasets by joining the different parts
```{r}
## We add SUBJID and VISIT again
db_female_knn_imputed <- bind_cols(f_nonnum, f_knn)
db_male_knn_imputed <- bind_cols(m_nonnum, m_knn)

## Add non-imputable columns
extra_female <- db_female %>%
  select(DEATHAGE, EVERSMOKE, HXSMOKE, PASTALC, ALCFREQ, EX3XWK, REGEX,
         EX_COMP10YRS, SLEEPING_HRS, NAPPING_HRS, SITTING_HRS)

extra_male <- db_male %>%
  select(DEATHAGE, EVERSMOKE, HXSMOKE, PASTALC, ALCFREQ, EX3XWK, REGEX,
         EX_COMP10YRS, SLEEPING_HRS, NAPPING_HRS, SITTING_HRS)

## Join everything
db_female_final <- bind_cols(db_female_knn_imputed, extra_female)
db_male_final <- bind_cols(db_male_knn_imputed, extra_male)
```


### Statistics on imputation per individual
```{r}
# For females
# Count NAs BEFORE imputation per individual across all visits
na_before_female <- db_female_final %>%
  group_by(SUBJID) %>%
  summarise(total_na_before = sum(across(everything(), ~sum(is.na(.))))) %>%
  pull(total_na_before)

# Count NAs AFTER imputation per individual across all visits
na_after_female <- db_female_knn_imputed %>%
  group_by(SUBJID) %>%
  summarise(total_na_after = sum(across(everything(), ~sum(is.na(.))))) %>%
  pull(total_na_after)

# Calculate number of imputed values per individual
n_imputed_female <- na_before_female - na_after_female

imputation_summary_female <- data.frame(
  n_imputed_vars = n_imputed_female
) %>%
  group_by(n_imputed_vars) %>%
  summarise(n_individuals = n()) %>%
  mutate(
    percentage = round(n_individuals / sum(n_individuals) * 100, 2),
    cumulative_pct = round(cumsum(n_individuals) / sum(n_individuals) * 100, 2)
  )

print("Imputation statistics - Females:")
print(imputation_summary_female)

# Individuals with no imputed data
no_imputation_female <- sum(n_imputed_female == 0)
print(paste("Total individuals (females):", length(n_imputed_female)))
print(paste("Individuals with NO imputed data (females):", no_imputation_female))

# Individuals with at least one imputed value
some_imputation_female <- sum(n_imputed_female > 0)
print(paste("Individuals with at least ONE imputed value (females):", some_imputation_female))

# Individuals with more than 50% of variables imputed (across all visits)
total_vars_female <- (ncol(db_female_final) - 2) * length(unique(db_female_final$VISIT))
majority_imputed_female <- sum(n_imputed_female > (total_vars_female / 2))
print(paste("Individuals with >50% of total variables imputed (females):", majority_imputed_female))

# Mean and median of imputed variables per individual
print(paste("Mean imputed variables per individual (females):", round(mean(n_imputed_female), 2)))
print(paste("Median imputed variables per individual (females):", median(n_imputed_female)))
print(paste("Min imputed variables (females):", min(n_imputed_female)))
print(paste("Max imputed variables (females):", max(n_imputed_female)))


# For males
# Count NAs BEFORE imputation per individual across all visits
na_before_male <- db_male_final %>%
  group_by(SUBJID) %>%
  summarise(total_na_before = sum(across(everything(), ~sum(is.na(.))))) %>%
  pull(total_na_before)

# Count NAs AFTER imputation per individual across all visits
na_after_male <- db_male_knn_imputed %>%
  group_by(SUBJID) %>%
  summarise(total_na_after = sum(across(everything(), ~sum(is.na(.))))) %>%
  pull(total_na_after)

# Calculate number of imputed values per individual
n_imputed_male <- na_before_male - na_after_male

imputation_summary_male <- data.frame(
  n_imputed_vars = n_imputed_male
) %>%
  group_by(n_imputed_vars) %>%
  summarise(n_individuals = n()) %>%
  mutate(
    percentage = round(n_individuals / sum(n_individuals) * 100, 2),
    cumulative_pct = round(cumsum(n_individuals) / sum(n_individuals) * 100, 2)
  )

print("Imputation statistics - Males:")
print(imputation_summary_male)

# Individuals with no imputed data
no_imputation_male <- sum(n_imputed_male == 0)
print(paste("Total individuals (males):", length(n_imputed_male)))
print(paste("Individuals with NO imputed data (males):", no_imputation_male))

# Individuals with at least one imputed value
some_imputation_male <- sum(n_imputed_male > 0)
print(paste("Individuals with at least ONE imputed value (males):", some_imputation_male))

# Individuals with more than 50% of variables imputed (across all visits)
total_vars_male <- (ncol(db_male_final) - 2) * length(unique(db_male_final$VISIT))
majority_imputed_male <- sum(n_imputed_male > (total_vars_male / 2))
print(paste("Individuals with >50% of total variables imputed (males):", majority_imputed_male))

# Mean and median of imputed variables per individual
print(paste("Mean imputed variables per individual (males):", round(mean(n_imputed_male), 2)))
print(paste("Median imputed variables per individual (males):", median(n_imputed_male)))
print(paste("Min imputed variables (males):", min(n_imputed_male)))
print(paste("Max imputed variables (males):", max(n_imputed_male)))

# Histogram for females
ggplot(data.frame(n_imputed = n_imputed_female), aes(x = n_imputed)) +
  geom_histogram(binwidth = 1, fill = "coral", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Imputed Variables per Individual - Females",
       x = "Number of Imputed Variables (across all visits)",
       y = "Number of Individuals") +
  theme_minimal()

# Histogram for males
ggplot(data.frame(n_imputed = n_imputed_male), aes(x = n_imputed)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Imputed Variables per Individual - Males",
       x = "Number of Imputed Variables (across all visits)",
       y = "Number of Individuals") +
  theme_minimal()
```

