---
title: "Clusters de variables Female separated hologic - other variables"
author: "Adriana Ibáñez"
date: "2025-10-29"
output: pdf_document
---

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(cluster)
library(ggplot2)
library(circlize)
library(tibble)

```

## Summary
This script performs clustering analysis on female variables based on their change patterns over multiple visits. It reads in data files, reshapes the data, calculates average values, computes crude derivatives, and applies clustering techniques to identify groups of variables with similar trajectories. The results are visualized through various plots and summarized in tables.

```{r}
# File routes
base_dir <- "dataframes_by_var"
female_dir <- file.path(base_dir, "female")
female_files <- list.files(female_dir, pattern = "\\.rds$", full.names = TRUE)
names(female_files) <- tools::file_path_sans_ext(basename(female_files))
```

```{r}
cat("Loading women data...\n")
data_list <- map(female_files, readRDS)
cat("Loaded variables:", length(data_list), "\n")
```

```{r}
cat("Construyendo tabla larga...\n")

long_data <- map_df(names(data_list), ~{
  df <- data_list[[.x]]
  var_name <- .x  # ex: "BMI_female"
  df %>%
    rownames_to_column("SUBJID") %>%
    pivot_longer(cols = starts_with("V"), names_to = "VISIT", values_to = "Value") %>%
    mutate(Variable = var_name, VISIT = as.numeric(sub("V", "", VISIT))) %>%
    select(SUBJID, Variable, VISIT, Value)
}, .id = NULL)

```

### Added for editing the data to remove problematic individual or variables

# After creating long_data, add this filter to remove the specific individual
long_data <- long_data %>%
  filter(SUBJID != "0ILJ00464")

# After creating long_data, add this filter to remove the specific variables
long_data <- long_data %>%
  filter(!Variable %in% c("SUBT_LBMC_female", "SUBT_LEAN_female", "TOT_LBMC_female", "TOT_LEAN_female", "TR_LBMC_female", "TR_LEAN_female", "SUBT_FAT_female", "TOT_FAT_female", "TESTOSTERONE_female", "TOT_TOT_female", "SUBT_TOT_female", "TR_TOT_female"))


```{r}
# Define the lists of variables for each group, adding "_female"
group_p_vars <- c("BIO_RES_female", "FATPERC_female", "FATMASS_female", "FATFREEMASS_female", "FATFREEPERC_female", "LEAN_FAT_female", "RATIO_LOWER_female", "RATIO_UPPER_female", "BODYH20_TOT_female", "BODYH20_PERC_female", "REST_ENERGY_female", "TARGETWT_LOWER_female", "TARGETWT_UPPER_female", "RECFAT_LOWER_female", "RECFAT_UPPER_female", "BIO_REA_female", "DCN_VERYLGHT_female", "DCN_LIGHT_female", "DCN_MOD_female", "DCN_HEAVY_female", "DCN_VERYHVY_female", "HD_AREA_female", "LA_AREA_female", "RA_AREA_female", "LRIB_AREA_female", "RRIB_AREA_female", "TSPI_AREA_female", "LSPI_AREA_female", "PELV_AREA_female", "LL_AREA_female", "RL_AREA_female", "SUBT_AREA_female", "TOT_AREA_female", "HD_BMC_female", "LA_BMC_female", "RA_BMC_female", "LRIB_BMC_female", "RRIB_BMC_female", "TSPI_BMC_female", "LSPI_BMC_female", "PELV_BMC_female", "LL_BMC_female", "RL_BMC_female", "TOT_BMC_female", "TR_BMC_female", "SUBT_BMC_female", "HD_BMD_female", "LA_BMD_female", "RA_BMD_female", "LRIB_BMD_female", "RRIB_BMD_female", "TSPI_BMD_female", "LSPI_BMD_female", "PELV_BMD_female", "LL_BMD_female", "RL_BMD_female", "SUBT_BMD_female", "TOT_BMD_female", "LA_FAT_female", "RA_FAT_female", "TR_FAT_female", "LL_FAT_female", "RL_FAT_female", "SUBT_FAT_female", "HD_FAT_female", "TOT_FAT_female", "LA_LEAN_female", "RA_LEAN_female", "TR_LEAN_female", "LL_LEAN_female", "RL_LEAN_female", "SUBT_LEAN_female", "HD_LEAN_female", "TOT_LEAN_female", "LA_LBMC_female", "RA_LBMC_female", "TR_LBMC_female", "LL_LBMC_female", "RL_LBMC_female", "SUBT_LBMC_female", "HD_LBMC_female", "TOT_LBMC_female", "LA_TOT_female", "RA_TOT_female", "TR_TOT_female", "LL_TOT_female", "RL_TOT_female", "SUBT_TOT_female", "HD_TOT_female", "TOT_TOT_female", "LA_PCN_female", "RA_PCN_female", "TR_PCN_female", "LL_PCN_female", "RL_PCN_female", "SUBT_PCN_female", "HD_PCN_female", "TOT_PCN_female")


# Create subsets of long_data for each group
# First, get all unique variables
all_vars <- unique(long_data$Variable)

# Group P: only those that exist in the data
group_p_vars <- group_p_vars[group_p_vars %in% all_vars]
long_data_p <- long_data %>% filter(Variable %in% group_p_vars)

# Group B: the rest
group_b_vars <- all_vars[!all_vars %in% c(group_p_vars)]
long_data_b <- long_data %>% filter(Variable %in% group_b_vars)

saveRDS(group_p_vars, file = "group_p_vars_female")
saveRDS(group_b_vars, file = "group_b_vars_female")
```


# PHYSICAL: 1 DERIVATIVE -> MEDIAN
```{r}
# 1. Calculate individual derivatives
wide_indiv_p <- long_data_p %>%
  select(SUBJID, Variable, VISIT, Value) %>%
  pivot_wider(names_from = VISIT, names_prefix = "V", values_from = Value)

deltas_indiv_p <- wide_indiv_p %>%
  mutate(
    dV8_V7  = V8 - V7,
    dV9_V8  = V9 - V8,
    dV10_V9 = V10 - V9
  ) %>%
  select(SUBJID, Variable, dV8_V7, dV9_V8, dV10_V9) %>%
  pivot_longer(
    cols = starts_with("dV"),
    names_to = "Intervalo",
    values_to = "Delta_indiv"
  ) %>%
  drop_na(Delta_indiv)

# 2. MEDIAN of individual change by variable and interval
deltas_summary_p <- deltas_indiv_p %>%
  group_by(Variable, Intervalo) %>%
  summarise(
    Median_Delta = median(Delta_indiv, na.rm = TRUE),
    Mean_Delta   = mean(Delta_indiv, na.rm = TRUE),
    SD_Delta     = sd(Delta_indiv, na.rm = TRUE),
    n_pairs      = n(),
    .groups = "drop"
  ) %>%
  filter(n_pairs >= 8)

# 3. Matrix for clustering
deltas_p <- deltas_summary_p %>%
  select(Variable, Intervalo, Median_Delta) %>%
  pivot_wider(names_from = Intervalo, values_from = Median_Delta) %>%
  drop_na()

deltas_matrix_p <- deltas_p %>%
  column_to_rownames("Variable") %>%
  as.matrix()

row_sds_p <- apply(deltas_matrix_p, 1, sd, na.rm = TRUE)
problematic_vars_p <- names(row_sds_p)[row_sds_p == 0 | is.na(row_sds_p)]
print(problematic_vars_p)

deltas_clean_p <- deltas_p %>%
  filter(!Variable %in% problematic_vars_p)


deltas_matrix_clean_p <- deltas_clean_p %>%
  column_to_rownames("Variable") %>%
  as.matrix()

# 4. Clustering (same as before)
diss_deltas_p <- dist(deltas_matrix_clean_p, method = "euclidean")


k_range <- 2:15
sil_scores_p <- numeric(length(k_range))

for (i in seq_along(k_range)) {
  k <- k_range[i]
  pam_fit <- pam(diss_deltas_p, k = k, diss = TRUE)
  sil <- silhouette(pam_fit$clustering, diss_deltas_p)
  sil_scores_p[i] <- mean(sil[, 3])
}

optimal_k_p <- k_range[which.max(sil_scores_p)]

sil_df_p <- data.frame(
  k = k_range,
  Silhouette_Score = round(sil_scores_p, 3),
  Optimal = ifelse(k_range == optimal_k_p, "Sí", "")
)

cat("SILHOUETTE SCORES for k (Group P Mujeres)\n")
print(sil_df_p)

p_sil_p <- ggplot(sil_df_p, aes(x = k, y = Silhouette_Score)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(size = 3, color = "steelblue") +
  geom_point(data = subset(sil_df_p, Optimal == "Sí"), color = "red", size = 4, shape = 17) +
  scale_x_continuous(breaks = k_range) +
  labs(
    title = "Silhouette Score - Women Physical Vars",
    subtitle = paste("optimal k =", optimal_k_p),
    x = "Clusters number (k)",
    y = "Average Silhouette"
  ) +
  theme_minimal(base_size = 15) +
  theme(plot.title = element_text(face = "bold"), plot.subtitle = element_text(color = "gray40")) +
  geom_text(aes(label = Silhouette_Score), vjust = -1.2, size = 3.5, color = "black")

print(p_sil_p)
ggsave(file.path(base_dir, "silhouette_scores_robust_group_p_female.png"), p_sil_p, width = 8, height = 5, dpi = 300)

# FINAL CLUSTERING
optimal_k_p <- 4
pam_deltas_p <- pam(diss_deltas_p, k = optimal_k_p, diss = TRUE)
deltas_clean_p$Cluster <- as.factor(pam_deltas_p$clustering)

# GRAPHS
plot_deltas_p <- deltas_clean_p %>%
  pivot_longer(
    cols = starts_with("d"),
    names_to = "Intervalo",
    values_to = "Cambio"
  ) %>%
  mutate(
    Intervalo = factor(Intervalo, levels = c("dV8_V7", "dV9_V8", "dV10_V9")),
    Intervalo = recode(Intervalo,
                       "dV8_V7" = "V8 - V7",
                       "dV9_V8" = "V9 - V8",
                       "dV10_V9" = "V10 - V9")
  )

p_deriv_p <- ggplot(plot_deltas_p, aes(x = Intervalo, y = Cambio, group = Variable, color = Cluster)) +
  geom_line(alpha = 0.6, size = 0.8) +
  geom_point(size = 1.5, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = paste("Patterns of change - Women Physical Vars (k =", optimal_k_p, ")"),
    subtitle = "Median of individual changes",
    x = "Interval between visits",
    y = "Medium change (original units)",
    color = "Cluster"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

print(p_deriv_p)
ggsave(file.path(base_dir, "derivadas_por_cluster_group_p_female.png"), p_deriv_p, width = 10, height = 6, dpi = 300)

# FACETED GRAPH
p_facet_p <- ggplot(plot_deltas_p, aes(x = Intervalo, y = Cambio, group = Variable)) +
  geom_line(alpha = 0.7, size = 0.9, color = "steelblue") +
  geom_point(size = 2, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  facet_wrap(~ Cluster, ncol = 2, labeller = labeller(Cluster = function(x) paste("Cluster", x))) +
  labs(
    title = paste("Patterns of change by cluster (k =", optimal_k_p, ") - Physical Women"),
    subtitle = "Robust method: median of individual changes",
    x = "Interval between visits",
    y = "Medium change (original units)"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40"),
    strip.text = element_text(face = "bold", size = 12),
    panel.grid.minor = element_blank()
  )

print(p_facet_p)
ggsave(file.path(base_dir, "derivadas_facet_group_p_female.png"), p_facet_p, width = 12, height = 8, dpi = 300)

# TABLA
tabla_clusters_p <- deltas_clean_p %>%
  count(Cluster, name = "N_Variables") %>%
  arrange(desc(N_Variables)) %>%
  mutate(
    Porcentaje = round(N_Variables / sum(N_Variables) * 100, 1),
    Ejemplos = map_chr(Cluster, ~ {
      vars <- deltas_clean_p %>% filter(Cluster == .x) %>% pull(Variable)
      paste(head(sort(vars), 3), collapse = ", ")
    })
  ) %>%
  select(Cluster, N_Variables, Porcentaje, Ejemplos)

cat("=== VARIABLES POR CLUSTER (método robusto, Group P Mujeres) ===\n")
print(tabla_clusters_p)
write.table(tabla_clusters_p, file.path(base_dir, "tabla_clusters_group_p_female.txt"),
            sep = "\t", row.names = FALSE, quote = FALSE)
```
## *Subclustering cluster 1 - Physical vars Women*

```{r}
# 1. Filter only the variables in Cluster 1
vars_cluster1 <- deltas_clean_p %>%
  filter(Cluster == 1) %>%
  pull(Variable)

cat("Número de variables en Cluster 1:", length(vars_cluster1), "\n")
cat("Ejemplos:", paste(head(sort(vars_cluster1), 10), collapse = ", "), "\n\n")

# 2. Recreate the array using only those variables
deltas_sub <- deltas_summary_p %>%                     # We used the original summary (it has all the variables)
  filter(Variable %in% vars_cluster1) %>%
  select(Variable, Intervalo, Median_Delta) %>%
  pivot_wider(names_from = Intervalo, values_from = Median_Delta) %>%
  drop_na()

deltas_matrix_sub <- deltas_sub %>%
  column_to_rownames("Variable") %>%
  as.matrix()

# Remove rows with zero variance (in case any remained constant within the subgroup)
row_sds_sub <- apply(deltas_matrix_sub, 1, sd, na.rm = TRUE)
problematic_sub <- names(row_sds_sub)[row_sds_sub == 0 | is.na(row_sds_sub)]

if(length(problematic_sub) > 0){
  cat("Variables eliminated due to zero variance in sub-cluster:", problematic_sub, "\n")
  deltas_matrix_sub <- deltas_matrix_sub[!rownames(deltas_matrix_sub) %in% problematic_sub, , drop = FALSE]
}

# 3. Distance and search for the optimal k (silhouette) for the sub-cluster
diss_sub <- dist(deltas_matrix_sub, method = "euclidean")

k_range_sub <- 2:15
sil_scores_sub <- numeric(length(k_range_sub))

for (i in seq_along(k_range_sub)) {
  k <- k_range_sub[i]
  pam_sub <- pam(diss_sub, k = k, diss = TRUE)
  sil <- silhouette(pam_sub$clustering, diss_sub)
  sil_scores_sub[i] <- mean(sil[, 3])
}

optimal_k_sub <- 4

cat("optimal k for sub-clustering of Cluster 1=", optimal_k_sub, "\n")

# Silhouette chart for the sub-cluster
sil_df_sub <- data.frame(
  k = k_range_sub,
  Silhouette_Score = round(sil_scores_sub, 3),
  Optimal = ifelse(k_range_sub == optimal_k_sub, "Sí", "")
)

p_sil_sub <- ggplot(sil_df_sub, aes(x = k, y = Silhouette_Score)) +
  geom_line(color = "darkorange", size = 1) +
  geom_point(size = 3, color = "darkorange") +
  geom_point(data = subset(sil_df_sub, Optimal == "Sí"), color = "red", size = 4, shape = 17) +
  scale_x_continuous(breaks = k_range_sub) +
  labs(title = "Silhouette Score - Sub-clustering Cluster 1 (Women Physical)",
       subtitle = paste("optimal k =", optimal_k_sub),
       x = "Number of sub-clusters", y = "Average Silhouette") +
  theme_minimal(base_size = 14) +
  geom_text(aes(label = Silhouette_Score), vjust = -1.2, size = 3.5)

print(p_sil_sub)
ggsave(file.path(base_dir, "silhouette_subcluster1_group_p_female.png"), p_sil_sub, width = 8, height = 5, dpi = 300)

# 4. Final clustering of the sub-cluster
pam_sub <- pam(diss_sub, k = optimal_k_sub, diss = TRUE)

deltas_sub$SubCluster <- as.factor(pam_sub$clustering)

# 5. Sub-clustering charts
plot_sub <- deltas_sub %>%
  pivot_longer(cols = starts_with("dV"),
               names_to = "Intervalo",
               values_to = "Cambio") %>%
  mutate(Intervalo = factor(Intervalo, levels = c("dV8_V7", "dV9_V8", "dV10_V9")),
         Intervalo = recode(Intervalo,
                            "dV8_V7" = "V8 - V7",
                            "dV9_V8" = "V9 - V8",
                            "dV10_V9" = "V10 - V9"))

# Colored lines by sub-cluster
p_sub1 <- ggplot(plot_sub, aes(x = Intervalo, y = Cambio, group = Variable, color = SubCluster)) +
  geom_line(alpha = 0.7, size = 0.9) +
  geom_point(size = 1.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  labs(title = paste("Sub-patterns inside Cluster 1 (k =", optimal_k_sub, ") - Women Physical"),
       subtitle = "Median of individual changes",
       x = "Interval between visits", y = "Medium change", color = "Sub-cluster") +
  theme_minimal(base_size = 15) +
  theme(legend.position = "bottom")

print(p_sub1)
ggsave(file.path(base_dir, "subcluster1_lines_group_p_female.png"), p_sub1, width = 11, height = 7, dpi = 300)

# Faceted by sub-cluster
p_facet_sub <- ggplot(plot_sub, aes(x = Intervalo, y = Cambio, group = Variable)) +
  geom_line(alpha = 0.7, color = "steelblue") +
  geom_point(size = 2, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  facet_wrap(~ SubCluster, ncol = 2, labeller = labeller(SubCluster = function(x) paste("SubCluster", x))) +
  labs(title = paste("Detailed trajectories - Cluster 1 subdivided (k =", optimal_k_sub, ")"),
       subtitle = "Women Physical variables - robust method") +
  theme_minimal(base_size = 15) +
  theme(strip.text = element_text(face = "bold"))

print(p_facet_sub)
ggsave(file.path(base_dir, "subcluster1_facet_group_p_female.png"), p_facet_sub, width = 12, height = 8, dpi = 300)

# 6. Summary table of sub-clusters
tabla_subclusters <- deltas_sub %>%
  count(SubCluster, name = "N_Variables") %>%
  arrange(desc(N_Variables)) %>%
  mutate(Porcentaje = round(N_Variables / sum(N_Variables) * 100, 1),
         Ejemplos = map_chr(SubCluster, ~{
           vars <- deltas_sub %>% filter(SubCluster == .x) %>% pull(Variable)
           paste(head(sort(vars), 4), collapse = ", ")
         }))

cat("SUB-CLUSTERS INSIDE CLUSTER 1 (Mujeres físicas) \n")
print(tabla_subclusters)
write.table(tabla_subclusters, file.path(base_dir, "tabla_subclusters_cluster1_group_p_female.txt"),
            sep = "\t", row.names = FALSE, quote = FALSE)
```
```{r}
vars_no_change1 <- deltas_sub$Variable[deltas_sub$SubCluster == 1]

saveRDS(vars_no_change1, "variables_no_cambio_subcluster_physical_female.rds")
```






## BIOCHEMICAL: 1ST DERIVATIVE -> MEDIAN

```{r}
# 1. Calculate individual derivatives
wide_indiv_b <- long_data_b %>%
  select(SUBJID, Variable, VISIT, Value) %>%
  pivot_wider(names_from = VISIT, names_prefix = "V", values_from = Value)

deltas_indiv_b <- wide_indiv_b %>%
  mutate(
    dV8_V7  = V8 - V7,
    dV9_V8  = V9 - V8,
    dV10_V9 = V10 - V9
  ) %>%
  select(SUBJID, Variable, dV8_V7, dV9_V8, dV10_V9) %>%
  pivot_longer(
    cols = starts_with("dV"),
    names_to = "Intervalo",
    values_to = "Delta_indiv"
  ) %>%
  drop_na(Delta_indiv)

# 2. MEDIAN of individual change
deltas_summary_b <- deltas_indiv_b %>%
  group_by(Variable, Intervalo) %>%
  summarise(
    Median_Delta = median(Delta_indiv, na.rm = TRUE),
    Mean_Delta   = mean(Delta_indiv, na.rm = TRUE),
    SD_Delta     = sd(Delta_indiv, na.rm = TRUE),
    n_pairs      = n(),
    .groups = "drop"
  ) %>%
  filter(n_pairs >= 8)

# 3. Matrix for clustering
deltas_b <- deltas_summary_b %>%
  select(Variable, Intervalo, Median_Delta) %>%
  pivot_wider(names_from = Intervalo, values_from = Median_Delta) %>%
  drop_na()

deltas_matrix_b <- deltas_b %>%
  column_to_rownames("Variable") %>%
  as.matrix()

row_sds_b <- apply(deltas_matrix_b, 1, sd, na.rm = TRUE)
problematic_vars_b <- names(row_sds_b)[row_sds_b == 0 | is.na(row_sds_b)]
cat("Variables with constant change (or NA) (Group B Mujeres):", length(problematic_vars_b), "\n")
print(problematic_vars_b)

deltas_clean_b <- deltas_b %>%
  filter(!Variable %in% problematic_vars_b)


deltas_matrix_clean_b <- deltas_clean_b %>%
  column_to_rownames("Variable") %>%
  as.matrix()

# 4. Clustering
diss_deltas_b <- dist(deltas_matrix_clean_b, method = "euclidean")


k_range <- 2:15
sil_scores_b <- numeric(length(k_range))

for (i in seq_along(k_range)) {
  k <- k_range[i]
  pam_fit <- pam(diss_deltas_b, k = k, diss = TRUE)
  sil <- silhouette(pam_fit$clustering, diss_deltas_b)
  sil_scores_b[i] <- mean(sil[, 3])
}

optimal_k_b <- k_range[which.max(sil_scores_b)]

sil_df_b <- data.frame(
  k = k_range,
  Silhouette_Score = round(sil_scores_b, 3),
  Optimal = ifelse(k_range == optimal_k_b, "Sí", "")
)


print(sil_df_b)

p_sil_b <- ggplot(sil_df_b, aes(x = k, y = Silhouette_Score)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(size = 3, color = "steelblue") +
  geom_point(data = subset(sil_df_b, Optimal == "Sí"), color = "red", size = 4, shape = 17) +
  scale_x_continuous(breaks = k_range) +
  labs(
    title = "Silhouette Score - Mujeres Biochemical Vars (método robusto)",
    subtitle = paste("optimal k =", optimal_k_b),
    x = "Número de clusters (k)",
    y = "Silhouette medio"
  ) +
  theme_minimal(base_size = 15) +
  theme(plot.title = element_text(face = "bold"), plot.subtitle = element_text(color = "gray40")) +
  geom_text(aes(label = Silhouette_Score), vjust = -1.2, size = 3.5, color = "black")

print(p_sil_b)
ggsave(file.path(base_dir, "silhouette_scores_robust_group_b_female.png"), p_sil_b, width = 8, height = 5, dpi = 300)

# FINAL CLUSTERING
optimal_k_b <- 4 
pam_deltas_b <- pam(diss_deltas_b, k = optimal_k_b, diss = TRUE)
deltas_clean_b$Cluster <- as.factor(pam_deltas_b$clustering)

# GRAPHICS
plot_deltas_b <- deltas_clean_b %>%
  pivot_longer(
    cols = starts_with("d"),
    names_to = "Intervalo",
    values_to = "Cambio"
  ) %>%
  mutate(
    Intervalo = factor(Intervalo, levels = c("dV8_V7", "dV9_V8", "dV10_V9")),
    Intervalo = recode(Intervalo,
                       "dV8_V7" = "V8 - V7",
                       "dV9_V8" = "V9 - V8",
                       "dV10_V9" = "V10 - V9")
  )

p_deriv_b <- ggplot(plot_deltas_b, aes(x = Intervalo, y = Cambio, group = Variable, color = Cluster)) +
  geom_line(alpha = 0.6, size = 0.8) +
  geom_point(size = 1.5, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = paste("Patterns of change - Women Biochemical Vars (k =", optimal_k_b, ")"),
    subtitle = "Median of individual changes",
    x = "Interval between visits",
    y = "Medium change (original units)",
    color = "Cluster"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

print(p_deriv_b)
ggsave(file.path(base_dir, "derivadas_robustas_por_cluster_group_b_female.png"), p_deriv_b, width = 10, height = 6, dpi = 300)

# FACETED GRAPH
p_facet_b <- ggplot(plot_deltas_b, aes(x = Intervalo, y = Cambio, group = Variable)) +
  geom_line(alpha = 0.7, size = 0.9, color = "steelblue") +
  geom_point(size = 2, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  facet_wrap(~ Cluster, ncol = 3, labeller = labeller(Cluster = function(x) paste("Cluster", x))) +
  labs(
    title = paste("Patterns of change - Women biochemical (k =", optimal_k_b, ")"),
    subtitle = "Median of individual changes",
    x = "Interval between visits",
    y = "Medium change (original units)"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40"),
    strip.text = element_text(face = "bold", size = 12),
    panel.grid.minor = element_blank()
  )

print(p_facet_b)
ggsave(file.path(base_dir, "derivadas_robustas_facet_group_b_female.png"), p_facet_b, width = 14, height = 9, dpi = 300)

# TABLE
tabla_clusters_b <- deltas_clean_b %>%
  count(Cluster, name = "N_Variables") %>%
  arrange(desc(N_Variables)) %>%
  mutate(
    Porcentaje = round(N_Variables / sum(N_Variables) * 100, 1),
    Ejemplos = map_chr(Cluster, ~ {
      vars <- deltas_clean_b %>% filter(Cluster == .x) %>% pull(Variable)
      paste(head(sort(vars), 3), collapse = ", ")
    })
  ) %>%
  select(Cluster, N_Variables, Porcentaje, Ejemplos)

cat("VARIABLES FOR CLUSTER (Group B Mujeres) \n")
print(tabla_clusters_b)
write.table(tabla_clusters_b, file.path(base_dir, "tabla_clusters_robustos_group_b_female.txt"),
            sep = "\t", row.names = FALSE, quote = FALSE)
```

## *Subclustering cluster 1 - Biochemical vars Women*

```{r}
# 1. Extract the variables that are in Cluster 1
vars_cluster1_b <- deltas_clean_b %>%
  filter(Cluster == 1) %>%
  pull(Variable)

cat("Variables en Cluster 1:", length(vars_cluster1_b), "\n")
cat("Ejemplos:", paste(head(sort(vars_cluster1_b), 8), collapse = ", "), "\n\n")

vars_cluster1_b <- vars_cluster1_b[vars_cluster1_b != "LDL_female"]

# If Cluster 1 is empty or has very few variables, we will notify you.
if(length(vars_cluster1_b) < 3) {
  cat("ADVERTENCIA: El Cluster 1 tiene menos de 3 variables → no se hace sub-clustering\n")
} else {

  # 2. Rebuild the matrix using only those variables
  deltas_sub_b <- deltas_summary_b %>%
    filter(Variable %in% vars_cluster1_b) %>%
    select(Variable, Intervalo, Median_Delta) %>%
    pivot_wider(names_from = Intervalo, values_from = Median_Delta) %>%
    drop_na()

  deltas_matrix_sub_b <- deltas_sub_b %>%
    column_to_rownames("Variable") %>%
    as.matrix()

  # Remove variables that are now constants within the subgroup
  row_sds_sub_b <- apply(deltas_matrix_sub_b, 1, sd, na.rm = TRUE)
  problematic_sub_b <- names(row_sds_sub_b)[row_sds_sub_b == 0 | is.na(row_sds_sub_b)]

  if(length(problematic_sub_b) > 0){
    cat("Variables eliminadas por varianza 0 en sub-cluster:", problematic_sub_b, "\n")
    deltas_matrix_sub_b <- deltas_matrix_sub_b[!rownames(deltas_matrix_sub_b) %in% problematic_sub_b, , drop = FALSE]
  }

  # 3. Distance and search for the best k (silhouette)
  diss_sub_b <- dist(deltas_matrix_sub_b, method = "euclidean")

  # Reasonable range of k (maximum up to the number of variables - 1)
  k_range_sub <- 2:15
  sil_scores_sub_b <- numeric(length(k_range_sub))

  for (i in seq_along(k_range_sub)) {
    k <- k_range_sub[i]
    pam_sub <- pam(diss_sub_b, k = k, diss = TRUE)
    sil <- silhouette(pam_sub$clustering, diss_sub_b)
    sil_scores_sub_b[i] <- mean(sil[, 3])
  }

  optimal_k_sub_b <- 3

  cat("k óptimo para sub-clustering del Cluster 1 =", optimal_k_sub_b, "\n\n")

  # silhouette scores Table
  sil_df_sub_b <- data.frame(
    k = k_range_sub,
    Silhouette_Score = round(sil_scores_sub_b, 3),
    Optimal = ifelse(k_range_sub == optimal_k_sub_b, "Sí", "")
  )
  print(sil_df_sub_b)

  # Graph silhouette sub-cluster
  p_sil_sub_b <- ggplot(sil_df_sub_b, aes(x = k, y = Silhouette_Score)) +
    geom_line(color = "darkorange", size = 1.2) +
    geom_point(size = 3.5, color = "darkorange") +
    geom_point(data = subset(sil_df_sub_b, Optimal == "Sí"), color = "red", size = 5, shape = 17) +
    scale_x_continuous(breaks = k_range_sub) +
    labs(
      title = "Silhouette Score - Sub-clustering Cluster 1 (Bioquímicas Mujeres)",
      subtitle = paste("k óptimo =", optimal_k_sub_b, "|", nrow(deltas_matrix_sub_b), "variables"),
      x = "Número de sub-clusters",
      y = "Silhouette medio"
    ) +
    theme_minimal(base_size = 14) +
    geom_text(aes(label = Silhouette_Score), vjust = -1.3, size = 4)

  print(p_sil_sub_b)
  ggsave(file.path(base_dir, "silhouette_subcluster1_bioquimicas_female.png"),
         p_sil_sub_b, width = 9, height = 5.5, dpi = 300)

  # 4. Sub-cluster final Clustering  
  pam_sub_b <- pam(diss_sub_b, k = optimal_k_sub_b, diss = TRUE)
  deltas_sub_b$SubCluster <- as.factor(pam_sub_b$clustering)

  # 5. Sub-clustering graphs
  plot_sub_b <- deltas_sub_b %>%
    pivot_longer(cols = starts_with("dV"),
                 names_to = "Intervalo",
                 values_to = "Cambio") %>%
    mutate(
      Intervalo = factor(Intervalo, levels = c("dV8_V7", "dV9_V8", "dV10_V9")),
      Intervalo = recode(Intervalo,
                         "dV8_V7" = "V8 - V7",
                         "dV9_V8" = "V9 - V8",
                         "dV10_V9" = "V10 - V9")
    )

  # Colored lines by sub-cluster
  p_lines_sub_b <- ggplot(plot_sub_b, aes(x = Intervalo, y = Cambio, group = Variable, color = SubCluster)) +
    geom_line(alpha = 0.7, size = 0.9) +
    geom_point(size = 2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
      title = paste("Sub-patrones dentro del Cluster 1 Bioquímicas Mujeres (k =", optimal_k_sub_b, ")"),
      subtitle = "Mediana de cambios individuales",
      x = "Intervalo entre visitas",
      y = "Cambio mediano",
      color = "Sub-cluster"
    ) +
    theme_minimal(base_size = 15) +
    theme(legend.position = "bottom")

  print(p_lines_sub_b)
  ggsave(file.path(base_dir, "subcluster1_lines_bioquimicas_female.png"), p_lines_sub_b, width = 11, height = 7, dpi = 300)

  # Sub-cluster faceted
  p_facet_sub_b <- ggplot(plot_sub_b, aes(x = Intervalo, y = Cambio, group = Variable)) +
    geom_line(alpha = 0.7, size = 0.9, color = "steelblue") +
    geom_point(size = 2.2, color = "steelblue") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    facet_wrap(~ SubCluster, ncol = 3,
               labeller = labeller(SubCluster = function(x) paste("SubCluster", x))) +
    labs(
      title = paste("Trayectorias detalladas - Subdivisión Cluster 1 (k =", optimal_k_sub_b, ")"),
      subtitle = "Mujeres - Variables Bioquímicas"
    ) +
    theme_minimal(base_size = 15) +
    theme(strip.text = element_text(face = "bold", size = 13))

  print(p_facet_sub_b)
  ggsave(file.path(base_dir, "subcluster1_facet_bioquimicas_female.png"), p_facet_sub_b, width = 14, height = 9, dpi = 300)

  # 6. Summary table of sub-clusters
  tabla_sub_b <- deltas_sub_b %>%
    count(SubCluster, name = "N_Variables") %>%
    arrange(desc(N_Variables)) %>%
    mutate(
      Porcentaje = round(N_Variables / sum(N_Variables) * 100, 1),
      Ejemplos = map_chr(SubCluster, ~{
        vars <- deltas_sub_b %>% filter(SubCluster == .x) %>% pull(Variable)
        paste(head(sort(vars), 5), collapse = ", ")
      })
    )

  cat("SUB-CLUSTERS WITHIN CLUSTER 1 (Women Biochemical)\n")
  print(tabla_sub_b)
  write.table(tabla_sub_b,
              file.path(base_dir, "tabla_subclusters_cluster1_bioquimicas_female.txt"),
              sep = "\t", row.names = FALSE, quote = FALSE)

  cat("\nSub-clustering del Cluster 1 completado y guardado.\n")
}
```


```{r}
vars_no_change2 <- deltas_sub_b$Variable[deltas_sub_b$SubCluster == 2]

saveRDS(vars_no_change2, "variables_no_cambio_bioquim_female.rds")

```



# EXPLORATORY ANALYSIS
## Table with interesting variables
```{r}
variables_interes <- "TESTOSTERONE_female"

# Sub-table only with the variable TESTOSTERONE_female
subtabla_testo <- wide_data_b %>%
  filter(Variable %in% variables_interes) %>%
  select(SUBJID, Variable, ID, starts_with("V")) %>%
  arrange(Variable, SUBJID)


# Save
write.csv(subtabla_testo, 
          file = file.path(base_dir, "testosterone_female_data.csv"),
          row.names = FALSE)

cat("Sub-tabla guardada en:", base_dir, "\n")

```

```{r}
# Filter data for TESTOSTERONE_female
testo_data <- long_data %>%
  filter(Variable == "TESTOSTERONE_female")

# Plot: VISIT on x, Value on y, lines and points per SUBJID
p_testo <- ggplot(testo_data, aes(x = VISIT, y = Value, group = SUBJID)) +
  geom_line(alpha = 0.5, color = "steelblue") +
  geom_point(size = 1, alpha = 0.5, color = "steelblue") +
  labs(
    title = "Trajectories of TESTOSTERONE_female by Individual",
    x = "Visit",
    y = "TESTOSTERONE Value"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = unique(testo_data$VISIT))

print(p_testo)
ggsave(file.path(base_dir, "testosterone_trajectories.png"), p_testo, width = 10, height = 6, dpi = 300)
```


```{r}
# Show the first rows
print(head(subtabla_test, 10))

# Save
write.csv(subtabla_test, 
          file = file.path(base_dir, "Cluster2_data.csv"),
          row.names = FALSE)


cat("Sub-tabla guardada en:", base_dir, "\n")
```

